---
title: Starting with Data
teaching: 50
exercises: 30
output:
  unilur::tutorial_html_solution:
    toc: true
    toc_float: true
    toc_depth: 3
    suffix: ""
    theme: flatly
    highlight: kate
    number_sections: no
    number_subsections: no
---

```{r knitr_init, echo=FALSE, cache=FALSE, include=FALSE}
library(knitr)
library(unilur)

## Global options
options(max.print="90")
opts_chunk$set(echo=TRUE,
               cache=FALSE, #TRUE
               prompt=FALSE,
               tidy=FALSE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=90)
options(width = 90)

# boxes custom
#remotes::install_github("koncina/unilur")

knitr::opts_template$set(alert = list(box.title = "Watch out!",
                                      box.body = list(fill = "#ffe0d9", colour = "black"),
                                      box.header = list(fill = "#FFAD99", colour = "black"),
                                      box.collapse = NULL,
                                      box.icon = "fa-exclamation-triangle"))
knitr::opts_template$set(solution = list(box.title = "Solution",
                                         box.body = list(fill = "#e6f6e7", colour = "black"),
                                         box.header = list(fill = "#ace1af", colour = "black"),
                                         box.icon = "fa-check-square",
                                         box.collapse = TRUE))
knitr::opts_template$set(objectives = list(box.title = "Objectives",
                                            box.body = list(fill = "#bbe8f4", colour = "black"),
                                            box.header = list(fill = "#64c9e6", colour = "black"),
                                            box.icon = "fa-info-circle",
                                            box.collapse = NULL))
knitr::opts_template$set(questions = list(box.title = "Questions",
                                      box.body = list(fill = "#fff9dc", colour = "black"),
                                      box.header = list(fill = "#ffec8b", colour = "black"),
                                      box.icon = "fa-search",
                                      box.collapse = NULL))
knitr::opts_template$set(information = list(box.title = "Keypoints",
                                       box.body = list(fill = "#fff9dc", colour = "black"),
                                      box.header = list(fill = "#ffec8b", colour = "black"),
                                      box.icon = "fa-search",
                                      box.collapse = NULL))

```

```{block, opts.label = "objectives" }

- Describe what a data frame is.
- Load external data from a .csv file into a data frame.
- Summarize the contents of a data frame.
- Subset values from data frames.
- Describe the difference between a factor and a string.
- Convert between strings and factors.
- Reorder and rename factors.
- Change how character strings are handled in a data frame.
- Examine and change date formats.
```

```{block, opts.label = "questions" }

- What is a data.frame?
- How can I read a complete csv file into R?
- How can I get basic summary information about my dataset?
- How can I change the way R treats strings in my dataset?
- Why would I want strings to be treated differently?
- How are dates represented in R and how can I change the format?

```

The main dataset used in this tutorial is the Danish kongerække digitized by BA students in the Digital Methods class on the basis of online resources in Danmarkshistorien and kongeraekke URL. 

You can download the dataset [here](https://) or use R to download it in your current folder:
```{r, eval=FALSE}
# download the dataset
download.file("https://raw.githubusercontent.com/adivea/r-history/main/episodes/data/kings.csv", 
              destfile = "data/kings.csv")

```
Please refrain from opening the file, we will load and open it first in R, to prevent your operating system from changing the file's format 

## What are data frames and tibbles?

Data frames are the *de facto* data structure for tabular data in `R`, and what
we use for data processing, statistics, and plotting.

A data frame is the representation of data in the format of a table where the
columns are vectors that all have the same length. Data frames are analogous to
the more familiar spreadsheet in programs such as Excel, with one key difference.
Because columns are vectors,
each column must contain a single type of data (e.g., characters, integers,
factors). For example, here is a figure depicting a data frame comprising a
numeric, a character, and a logical vector.

![](fig/data-frame.svg){alt='A 3 by 3 data frame with columns showing numeric, character and logical values.'}

Data frames can be created by hand, but most commonly they are generated by the
functions `read_csv()` or `read_table()`; in other words, when importing
spreadsheets from your hard drive (or the web). We will now demonstrate how to
import tabular data using `read_csv()`.

## Presentation of the 'kings' data

We will be using a subset of the cleaned version of the dataset that
was produced through cleaning in OpenRefine (`data/kings.csv`). In this dataset, the missing data is encoded as "NULL", each row holds
information for a single interview respondent, and the columns
represent:

| column\_name          | description                                                                                                                      | 
| -------------------- | -------------------------------------------------------------------------------------------------------------------------------- |
| Name               | The unique name of Danish monarch   | 
| Start\_year              | starting year of the monarch's rule |
| End\_year              | the last year of the monarch's rule |
| Birth\_year              | the year the monarch was born |
| Death\_year              | the year the monarch died |
| House              | the royal house the monarch belonged to|
| Dynasty              | another term for the royal house|



## Importing data

You are going to load the data in R's memory using the function `read_csv()`
from the **`readr`** package, which is part of the **`tidyverse`**; learn
more about the **`tidyverse`** collection of packages
[here](https://www.tidyverse.org/).
**`readr`** gets installed as part as the **`tidyverse`** installation.
When you load  the **`tidyverse`** (`library(tidyverse)`), the core packages
(the packages used in most data analyses) get loaded, including **`readr`**.

Before proceeding, however, this is a good opportunity to talk about conflicts.
Certain packages we load can end up introducing function names that are already
in use by pre-loaded R packages. For instance, when we load the tidyverse 
package below, we will introduce two conflicting functions: `filter()` and `lag()`.
This happens because `filter` and `lag` are already functions used by the stats
package (already pre-loaded in R). What will happen now is that if we, for
example, call the `filter()` function, R will use the `dplyr::filter()` version
and not the `stats::filter()` one. This happens because, if conflicted, by 
default R uses the function from the most recently loaded package. Conflicted
functions may cause you some trouble in the future, so it is important that we
are aware of them so that we can properly handle them, if we want.

To do so, we just need the following functions from the conflicted package:  

- `conflicted::conflict_scout()`: Shows us any conflicted functions.  
- `conflict_prefer("function", "package_prefered")`:  Allows us to choose the
default function we want from now on.  

It is also important to know that we can, at any time, just call the function
directly from the package we want, such as `stats::filter()`.

Before we can use the `read_csv()` functions, we need to load the
tidyverse packages.

Also, if you recall, the missing data is encoded as "NULL" in the dataset.
We'll tell it to the function, so R will automatically convert all the "NULL"
entries in the dataset into `NA`.

```{r, eval=TRUE, message=FALSE, purl=FALSE}
library(tidyverse)

kings <- read_csv("data/kings.csv", na = c("NULL",""), n_max = 54)

```

In the above code, we notice the `read_csv()` function takes folder and file names
as inputs (e.g., `"data/kings.csv"`), each enclosed in quotations
(`""`) and separated by a comma.

If you were to type in the code above, it is likely that the `read.csv()`
function would appear in the automatically populated list of functions. This
function is different from the `read_csv()` function, as it is included in the
"base" packages that come pre-installed with R. Overall, `read.csv()` behaves
similar to `read_csv()`, with a few notable differences. First, `read.csv()`
coerces column names with spaces and/or special characters to different names. Second, `read.csv()`
stores data as a `data.frame`, where `read_csv()` stores data as a `tibble`.
We prefer tibbles because they have nice printing properties among other
desirable qualities. Read more about tibbles
[here](https://tibble.tidyverse.org/).

The second statement in the code above creates a data frame but doesn't output
any data because, as you might recall, assignments (`<-`) don't display
anything. (Note, however, that `read_csv` may show informational
text about the data frame that is created.) If we want to check that our data
has been loaded, we can see the contents of the data frame by typing its name:
`kings` in the console.

```{r, results="hold", purl=FALSE}
kings
## Try also
## view(kings)
## head(kings)
```


```{block, box.title = "Note", box.body = list(fill = "white"), box.icon = "fa-star"}
`read_csv()` assumes that fields are delimited by commas. However, in several
countries, the comma is used as a decimal separator and the semicolon (;) is
used as a field delimiter. If you want to read in this type of files in R,
you can use the `read_csv` function. It behaves exactly like `read_csv` but
uses different parameters for the decimal and the field separators. If you are
working with another format, they can be both specified by the user. Check out
the help for `read_csv()` by typing `?read_csv` to learn more. There is also
the `read_tsv()` for tab-separated data files, and `read_delim()` allows you
to specify more details about the structure of your file.
```


Note that `read_csv()` actually loads the data as a tibble.
A tibble is an extension of `R` data frames used by the **`tidyverse`**. When
the data is read using `read_csv()`, it is stored in an object of class
`tbl_df`, `tbl`,  and `data.frame`. You can see the class of an object with

```{r, results="hold", purl=FALSE}
class(kings)
class(kings$Start_year)
```

As a `tibble`, the type of data included in each column is listed in an
abbreviated fashion below the column names. For instance, here `Danish_kings` is a
column of characters (`<chr>`), 
`Start_year` and `End_year` are columns of floating point numbers (abbreviated `<dbl>` for the word 'double').

## Inspecting data frames

When calling a `tbl_df` object (like `kings` here), there is already a lot
of information about our data frame being displayed such as the number of rows,
the number of columns, the names of the columns, and as we just saw the class of
data stored in each column. However, there are functions to extract this
information from data frames.  Here is a non-exhaustive list of some of these
functions. Let's try them out!

Size:

- `dim(kings)` - returns a vector with the number of rows as the first
  element, and the number of columns as the second element (the **dim**ensions of
  the object)
- `nrow(kings)` - returns the number of rows
- `ncol(kings)` - returns the number of columns

Content:

- `head(kings)` - shows the first 6 rows
- `tail(kings)` - shows the last 6 rows

Names:

- `names(kings)` - returns the column names (synonym of `colnames()` for
  `data.frame` objects)

Summary:

- `str(kings)` - structure of the object and information about the class,
  length and content of each column
- `summary(kings)` - summary statistics for each column
- `glimpse(kings)` - returns the number of columns and rows of the tibble,
  the names and class of each column, and previews as many values will fit on the
  screen. Unlike the other inspecting functions listed above, `glimpse()` is not a
  "base R" function so you need to have the `dplyr` or `tibble` packages loaded to
  be able to execute it.

Note: most of these functions are "generic." They can be used on other types of
objects besides data frames or tibbles.

## Subsetting data frames

Our `kings` data frame has rows and columns (it has 2 dimensions).
In practice, we may not need the entire data frame; for instance, we may only
be interested in a subset of the observations (the rows) or a particular set
of variables (the columns). If we want to
access some specific data from it, we need to specify the "coordinates" (i.e., indices) we
want from it. Row numbers come first, followed by column numbers.


```{block, box.title = "Tip", box.body = list(fill = "white"), box.icon = "fa-star"}
Subsetting a `tibble` with `[` always results in a `tibble`.
However, note this is not true in general for data frames, so be careful!
Different ways of specifying these coordinates can lead to results with
different classes. This is covered in the Software Carpentry lesson
[R for Reproducible Scientific Analysis](https://swcarpentry.github.io/r-novice-gapminder/).
```

```{r, purl=FALSE}
## first element in the first column of the tibble
kings[1, 1]

## first element in the 6th column of the tibble 
kings[1, 6]

## first column of the tibble (as a vector)
kings[[1]]

## first column of the tibble
kings[1]

## first three elements in the 7th column of the tibble
kings[1:3, 7]

## the 3rd row of the tibble
kings[3, ]

## equivalent to head_kings <- head(kings)
head_kings <- kings[1:6, ]
```

`:` is a special function that creates numeric vectors of integers in increasing
or decreasing order, test `1:10` and `10:1` for instance.

You can also exclude certain indices of a data frame using the "`-`" sign:

```{r, purl=FALSE}
kings[, -1]          # The whole tibble, except the first column

kings[-c(7:54), ]   # Equivalent to head(kings)
```

`tibble`s can be subset by calling indices (as shown previously), but also by
calling their column names directly:

```{r, eval=FALSE, purl=FALSE}
kings["Name"]       # Result is a tibble

kings[, "Name"]     # Result is a tibble

kings[["Name"]]     # Result is a vector

kings$Name          # Result is a vector
```

In RStudio, you can use the autocompletion feature to get the full and correct
names of the columns.


### Exercise 1

1. Create a tibble (`kings_20`) containing only the data in
  row 20 of the `kings` dataset.

2. Notice how `nrow()` gave you the number of rows in the tibble?

- Use that number to pull out just that last row in the tibble.
- Compare that with what you see as the last row using `tail()` to make
  sure it's meeting expectations.
- Pull out that last row using `nrow()` instead of the row number.
- Create a new tibble (`kings_last`) from that last row.

3. Using the number of rows in the kings dataset that you found in
  question 2, extract the row that is in the middle of the dataset. Store
  the content of this middle row in an object named `kings_middle`.
  (hint: This dataset has an odd number of rows, so finding the middle is a
  bit trickier than dividing n\_rows by 2. Use the median( ) function and what
  you've learned about sequences in R to extract the middle row!

4. Combine `nrow()` with the `-` notation above to reproduce the behavior of
  `head(kings)`, keeping just the first through 6th rows of the
  kings dataset.

```{r, solution = TRUE }

## 1.
kings_20 <- kings[20, ]

## 2.
# Saving `n_rows` to improve readability and reduce duplication
n_rows <- nrow(kings)
kings_last <- kings[n_rows, ]

## 3.
kings_middle <- kings[round(median(1:n_rows)), ]

## 4.
kings_head <- kings[-(7:n_rows), ]

```


## Calculating a new column

Sometimes you wants to create derived data and attach them to the table in a new column. The number of reign years, for example can be calculated by subtracting the starting year from the ending year. The mid-year of the reign can be calculated by subtracting half of reign years from the ending year. 

There are two ways of creating new columns: we can use the `$` operator or the `mutate()` function. We will learn the latter in the next lesson, and focus on `$` approach here.

First, you define an empty column and then use the assignment key `<-` to write the product of the subtraction to it.

```{r reign-years}
kings$Reign_duration <- kings$End_year - kings$Start_year
```

Investigate the result visually. Can you find out what kind of datatype the new column is?

```{r}
kings # see the entire tibble including the new column
kings$Reign_duration  # see the data as a vector
```

Now, create another column called `Midyear` by figuring out what the middle year of each ruler's reign is. This will be important if we wish to plot the rulers on a graph's `x` axis.

```{r midyear}
kings$Midyear <- kings$End_year - (kings$Reign_duration/2)
```

## Factors

R has a special data class, called factor, to deal with categorical data that
you may encounter when creating plots or doing statistical analyses. Factors are
very useful and actually contribute to making R particularly well suited to
working with data. So we are going to spend a little time introducing them.

Factors represent categorical data. They are stored as integers associated with
labels and they can be ordered (ordinal) or unordered (nominal). Factors
create a structured  relation between the different levels (values) of a
categorical variable, such as days of the week or responses to a question in
a survey. This can make it easier to see how one element relates to the
other elements in a column. While factors look (and often behave) like
character vectors, they are actually treated as integer vectors by `R`. So
you need to be very careful when treating them as strings.

Once created, factors can only contain a pre-defined set of values, known as
*levels*. By default, R always sorts levels in alphabetical order. For
instance, if you have a factor with 2 levels:

```{r, purl=TRUE}
animals <- factor(c("horse", "cat", "cat", "dog"))
```

R will assign `1` to the level `"cat"` and `2` to the level `"dog"`
(because `c` comes before `d`, even though the first element in this vector is
`"horse"`). You can see this by using the function `levels()` and you can find
the number of levels using `nlevels()`:

```{r, purl=FALSE}
levels(animals)

nlevels(animals)
```

Sometimes, the order of the factors does not matter. Other times you might want
to specify the order because it is meaningful (e.g., "low", "medium", "high").
It may improve your visualization, or it may be required by a particular type of
analysis. Here, one way to reorder our levels in the `animals`
vector would be:

```{r, results=TRUE, purl=FALSE}
animals # current order

animals <- factor(animals, levels = c("horse", "cat"))

animals # after re-ordering
```

In R's memory, these factors are represented by integers (1, 2), but are more
informative than integers because factors are self describing: `"horse"`,
`"cat"` is more descriptive than `1`, and `2`. Which one is "cat"? You
wouldn't be able to tell just from the integer data. Factors, on the other hand,
have this information built in. It is particularly helpful when there are many
levels. It also makes renaming levels easier. Let's say we made a mistake and
need to recode "cat" to "Cat". We can do this using the `fct_recode()`
function from the **`forcats`** package (included in the **`tidyverse`**) which
provides some extra tools to work with factors.

```{r, results=TRUE, purl=FALSE}
levels(animals)

animals <- fct_recode(animals, cat = "Cat")

## as an alternative, we could change the "cat" level directly using the
## levels() function, but we have to remember that "cat" is the second level
# levels(animals)[2] <- "Cat"

levels(animals)

animals
```

So far, your factor is unordered, like a nominal variable. R does not know the
difference between a nominal and an ordinal variable. You make your factor an
ordered factor by using the `ordered=TRUE` option inside your factor function.
Note how the reported levels changed from the unordered factor above to the
ordered version below. Ordered levels use the less than sign `<` to denote
level ranking.

```{r, results=TRUE, purl=FALSE}
animals_ordered <- factor(animals, 
                                        ordered = TRUE)

animals_ordered # after setting as ordered factor
```

### Converting factors

If you need to convert a factor to a character vector, you use
`as.character(x)`.

```{r, purl=FALSE}
as.character(animals)
```

Converting factors where the levels appear as numbers (such as concentration
levels, or years) to a numeric vector is a little trickier. The `as.numeric()`
function returns the index values of the factor, not its levels, so it will
result in an entirely new (and unwanted in this case) set of numbers.
One method to avoid this is to convert factors to characters, and then to
numbers. Another method is to use the `levels()` function. Compare:

```{r, purl=TRUE}
year_fct <- factor(c(1990, 1983, 1977, 1998, 1990))

as.numeric(year_fct)                     # Wrong! And there is no warning...

as.numeric(as.character(year_fct))       # Works...

as.numeric(levels(year_fct))[year_fct]   # The recommended way.
```

Notice that in the recommended `levels()` approach, three important steps occur:

- We obtain all the factor levels using `levels(year_fct)`
- We convert these levels to numeric values using `as.numeric(levels(year_fct))`
- We then access these numeric values using the underlying integers of the
  vector `year_fct` inside the square brackets
  
### Renaming factors

When your data is stored as a factor, you can use the `plot()` function to get a
quick glance at the number of observations represented by each factor level.
Let's extract the `Gender` column from our data frame, convert it into a
factor, and use it to look at the number of men and women in different royal Genders:

```{r factor-plot-default-order, purl=TRUE}
## create a vector from the data frame column "Gender"
Gender <- kings$Gender

## convert it into a factor
Gender <- as.factor(Gender)

## let's see what it looks like
Gender

## bar plot of the male and female monarchs:
plot(Gender)
```

Looking at the plot compared to the output of the vector, we can see that in
addition to "no"s and "yes"s, there are some respondents for which the
information about whether they were part of an irrigation association hasn't
been recorded, and encoded as missing data. They do not appear on the plot.
Let's encode them differently so they can counted and visualized in our plot.

```{r factor-plot-reorder, fig.alt="Bar plot of association membership, showing missing responses.", purl=TRUE}
## Let's recreate the vector from the data frame column "Gender"
Gender <- kings$Gender

## replace the missing data with "undetermined"
Gender[is.na(Gender)] <- "undetermined"

## convert it into a factor
Gender <- as.factor(Gender)

## let's see what it looks like
Gender

## bar plot of the monarchs by Gender:
plot(Gender)
```

### Exercise 2

- Rename the levels of the factor to full words with the first letter in uppercase:
  "Male","Undetermined", and "Female".

- Now that we have renamed the factor level to "Undetermined", can you
  create a barplot such that "Undetermined" is last (after "Female")?


```{r, solution = TRUE}
## Rename levels.
Gender <- fct_recode(Gender, Male = "M",
                         Undetermined = "undetermined", Female = "F")
## Reorder levels. Note we need to use the new level names.
Gender <- factor(Gender, levels = c("Male", "Female", "Undetermined"))
plot(Gender)
```


## Formatting Dates

One of the most common issues that new (and experienced!) R users have is
converting date and time information into a variable that is appropriate and
usable during analyses. A best
practice for dealing with date data is to ensure that each component of your
date is available as a separate variable. 

We are going to use the package **`lubridate`**, which is included in the
**`tidyverse`** installation but not loaded by default, so we have to load
it explicitly with `library(lubridate)`.

Start by loading the required package:

```{r load-package, message=FALSE, purl=FALSE}
library(lubridate)
```

The lubridate function `ymd()` takes a vector representing year, month, and day,
and converts it to a `Date` vector. `Date` is a class of data recognized by R as
being a date and can be manipulated as such. The argument that the function
requires is flexible, but, as a best practice, is a character vector formatted
as "YYYY-MM-DD".

Let's say we have a vector of dates in character format:

```{r, purl=FALSE}
char_dates <- c("7/31/2012", "8/9/2014", "4/30/2016")
str(char_dates)
```

We can convert this vector to dates as :

```{r, purl=FALSE}
as_date(char_dates, format = "%m/%d/%Y")
```

Argument `format` tells the function the order to parse the characters and
identify the month, day and year. The format above is the equivalent of
mm/dd/yyyy. A wrong format can lead to parsing errors or incorrect results.

For example, observe what happens when we use a lower case y instead of upper
case Y for the year.

```{r, purl=FALSE}
as_date(char_dates, format = "%m/%d/%y")
```

Here, the `%y` part of the format stands for a two-digit year instead of a
four-digit year, and this leads to parsing errors.

Or in the following example, observe what happens when the month and day
elements of the format are switched.

```{r, purl=FALSE}
as_date(char_dates, format = "%d/%m/%y")
```

Since there is no month numbered 30 or 31, the first and third dates cannot be
parsed.

We can also use functions `ymd()`, `mdy()` or `dmy()` to convert character
variables to date.

```{r, purl=FALSE}
mdy(char_dates)
```


## Recap Exercises

Let us apply skills in this and previous lessons and make calculations on a column from the kings dataset.

1. Using the kings dataset, eliminate the NAs from the years of reign.

```{r, }
kings$Reign_duration
```

```{r, solution = TRUE}

kings$Reign_duration <- kings$End_year-kings$Start_year
years <- kings$Reign_duration
years_no_na <- years[!is.na(years)]
# or
years_no_na <- na.omit(years)
```

2. Use the function `mean()` to calculate what is the average reign duration.

```{r, solution = TRUE}
mean(years, na.rm = TRUE)
```


3. Use R to figure out which kings enjoyed more than average years of rule.

```{r, solution = TRUE }
years_above_20 <- years_no_na[years_no_na > 20]
length(years_above_20)

```




```{block, opts.label = "information"}

- Load external data from a .csv file into a data frame.
- Summarize the contents of a data frame.
- Subset values from data frames.
- Describe the difference between a factor and a string.
- Convert between strings and factors.
- Reorder and rename factors.
- Change how character strings are handled in a data frame.
- Examine and change date formats.
- Use `lubridate` package to fix dates in R
```


