---
title: Data Wrangling with dplyr
teaching: 25
exercises: 15
output:
  unilur::tutorial_html_solution:
    toc: true
    toc_float: true
    toc_depth: 3
    suffix: ""
    theme: flatly
    highlight: kate
    number_sections: no
    number_subsections: no
---

```{r knitr_init, echo=FALSE, cache=FALSE, include=FALSE}
library(knitr)
library(unilur)

## Global options
options(max.print="90")
opts_chunk$set(echo=TRUE,
               cache=FALSE, #TRUE
               prompt=FALSE,
               tidy=FALSE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=90)
options(width = 90)

# boxes custom
#remotes::install_github("koncina/unilur")

knitr::opts_template$set(alert = list(box.title = "Watch out!",
                                      box.body = list(fill = "#ffe0d9", colour = "black"),
                                      box.header = list(fill = "#FFAD99", colour = "black"),
                                      box.collapse = NULL,
                                      box.icon = "fa-exclamation-triangle"))
knitr::opts_template$set(solution = list(box.title = "Solution",
                                         box.body = list(fill = "#e6f6e7", colour = "black"),
                                         box.header = list(fill = "#ace1af", colour = "black"),
                                         box.icon = "fa-check-square",
                                         box.collapse = TRUE))
knitr::opts_template$set(objectives = list(box.title = "Objectives",
                                            box.body = list(fill = "#bbe8f4", colour = "black"),
                                            box.header = list(fill = "#64c9e6", colour = "black"),
                                            box.icon = "fa-info-circle",
                                            box.collapse = NULL))
knitr::opts_template$set(questions = list(box.title = "Questions",
                                      box.body = list(fill = "#fff9dc", colour = "black"),
                                      box.header = list(fill = "#ffec8b", colour = "black"),
                                      box.icon = "fa-search",
                                      box.collapse = NULL))
knitr::opts_template$set(information = list(box.title = "Keypoints",
                                       box.body = list(fill = "#fff9dc", colour = "black"),
                                      box.header = list(fill = "#ffec8b", colour = "black"),
                                      box.icon = "fa-search",
                                      box.collapse = NULL))

#source("data/download_data.R")
```

```{block, opts.label = "objectives" }

- Describe the purpose of an R package and the **`dplyr`** package.
- Select certain columns in a dataframe with the **`dplyr`** function `select`.
- Select certain rows in a dataframe according to filtering conditions with the **`dplyr`** function `filter`.
- Link the output of one **`dplyr`** function to the input of another function with the 'pipe' operator `%>%`.
- Add new columns to a dataframe that are functions of existing columns with `mutate`.
- Use the split-apply-combine concept for data analysis.
- Use `summarize`, `group_by`, and `count` to split a dataframe into groups of observations, apply a summary statistics for each group, and then combine the results.

```

```{block, opts.label = "questions" }

- How can I select specific rows and/or columns from a dataframe?
- How can I combine multiple commands into a single command?
- How can I create new columns or remove existing columns from a dataframe?

```


**`dplyr`** is a package for making tabular data wrangling easier by using a
limited set of functions that can be combined to extract and summarize insights
from your data.

Like **`readr`**, **`dplyr`** is a part of the tidyverse. These packages were loaded
in R's memory when we called `library(tidyverse)` earlier.


```{block, box.title = "Note", box.body = list(fill = "white"), box.icon = "fa-star"}

The packages in the tidyverse, namely **`dplyr`**, **`tidyr`** and **`ggplot2`**
accept both the British (e.g. *summarise*) and American (e.g. *summarize*) spelling
variants of different function and option names. For this lesson, we utilize
the American spellings of different functions; however, feel free to use
the regional variant for where you are teaching.

```

## What is an R package?

The package **`dplyr`** provides easy tools for the most common data
wrangling tasks. It is built to work directly with dataframes, with many
common tasks optimized by being written in a compiled language (C++) (not all R
packages are written in R!).

There are also packages available for a wide range of tasks including building plots
(**`ggplot2`**, which we'll see later), downloading data from the NCBI database, or
performing statistical analysis on your data set. Many packages such as these are
housed on, and downloadable from, the **C**omprehensive **R** **A**rchive **N**etwork
(CRAN) using `install.packages`. This function makes the package accessible by your R
installation with the command `library()`, as you did with `tidyverse` earlier.

To easily access the documentation for a package within R or RStudio, use
`help(package = "package_name")`.

To learn more about **`dplyr`** after the workshop, you may want to check out this
[handy data transformation with **`dplyr`** cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-transformation.pdf).



```{block, box.title = "Note", box.body = list(fill = "white"), box.icon = "fa-star"}
There are alternatives to the `tidyverse` packages for data wrangling, including
the package [`data.table`](https://rdatatable.gitlab.io/data.table/). See this
[comparison](https://mgimond.github.io/rug_2019_12/Index.html)
for example to get a sense of the differences between using `base`, `tidyverse`, and
`data.table`.

```

## Learning **`dplyr`**

To make sure everyone will use the same dataset for this lesson, we'll read
again the kings dataset that we downloaded earlier.

```{r, results="hide", purl=FALSE, message=FALSE}

## load the tidyverse
library(tidyverse)

## load the data
kings <- read_csv("data/kings.csv", na = "NULL")

## inspect the data
kings

## preview the data
# view(kings)
```

We're going to learn some of the most common **`dplyr`** functions:

- `select()`: subset columns
- `filter()`: subset rows on conditions
- `mutate()`: create new columns by using information from other columns
- `group_by()` and `summarize()`: create summary statistics on grouped data
- `arrange()`: sort results
- `count()`: count discrete values

## Selecting columns and filtering rows

To select columns of a dataframe, use `select()`. The first argument to this
function is the dataframe (`kings`), and the subsequent arguments are the
columns to keep, separated by commas. Alternatively, if you are selecting
columns adjacent to each other, you can use a `:` to select a range of columns,
read as "select columns from \_\_\_ to \_\_\_." You may have done something similar in
the past using subsetting. `select()` is essentially doing the same thing as
subsetting, using a package (`dplyr`) instead of R's base functions.

```{r, results="hide", purl=FALSE}
# to select columns throughout the dataframe
select(kings, Name, Start_year, End_year)
# to do the same thing with subsetting
kings[c("Name", "Start_year", "End_year")]
# to select a series of connected columns
select(kings, Name:End_year)
```

To choose rows based on specific criteria, we can use the `filter()` function.
The argument after the dataframe is the condition we want our final
dataframe to adhere to (e.g. House is Gorm):

```{r, purl=FALSE}
# filters observations where House is "Gorm"
filter(kings, House == "Gorm")
```

We can also specify multiple conditions within the `filter()` function. We can
combine conditions using either "and" or "or" statements. In an "and"
statement, an observation (row) must meet **every** criteria to be included
in the resulting dataframe. To form "and" statements within dplyr, we can  pass
our desired conditions as arguments in the `filter()` function, separated by
commas:

```{r, purl=FALSE}

# filters observations with "and" operator (comma)
# output dataframe satisfies ALL specified conditions
filter(kings, House == "Estridsen",
                   End_year > 1100,
                    End_year < 1210)
```

We can also form "and" statements with the `&` operator instead of commas:

```{r, purl=FALSE}
# filters observations with "&" logical operator
# output dataframe satisfies ALL specified conditions
filter(kings, House == "Estridsen" &
                   End_year > 1100 &
                    End_year < 1210)
```

In an "or" statement, observations must meet *at least one* of the specified conditions.
To form "or" statements we use the logical operator for "or," which is the vertical bar (|):

```{r, purl=FALSE}
# filters observations with "|" logical operator
# output dataframe satisfies AT LEAST ONE of the specified conditions
filter(kings, House == "Estridsen" | End_year < 1100)
```

## Pipes

What if you want to select and filter at the same time? There are three
ways to do this: use intermediate steps, nested functions, or pipes.

With intermediate steps, you create a temporary dataframe and use
that as input to the next function, like this:

```{r, purl=FALSE}
kings_male <- filter(kings, Gender == "M")
kings_new <- select(kings_male, Name:Death_year)
kings_new
```

This is readable, but can clutter up your workspace with lots of objects that
you have to name individually. With multiple steps, that can be hard to keep
track of.

You can also nest functions (i.e. one function inside of another), like this:

```{r, purl=FALSE}
kings_new <- select(filter(kings, House == "Estridsen"),
                         Name:End_year)
```

This is handy, but can be difficult to read if too many functions are nested, as
R evaluates the expression from the inside out (in this case, filtering, then
selecting).

The last option, *pipes*, are a recent addition to R. Pipes let you take the
output of one function and send it directly to the next, which is useful when
you need to do many things to the same dataset. Pipes in R look like `%>%` and
are made available via the **`magrittr`** package, installed automatically with
**`dplyr`**. If you use RStudio, you can type the pipe with:

- <kbd>Ctrl</kbd> + <kbd>Shift</kbd> + <kbd>M</kbd> if you have a PC or <kbd>Cmd</kbd> +
  <kbd>Shift</kbd> + <kbd>M</kbd> if you have a Mac.

```{r, purl=FALSE}
kings %>%
    filter(House == "Gorm") %>%
    select(Name:End_year)
```

In the above code, we use the pipe to send the `kings` dataset first
through `filter()` to keep rows where `House` is "Gorm", then through
`select()` to keep only the columns from `Name` to `End_year`. Since `%>%`
takes the object on its left and passes it as the first argument to the function
on its right, we don't need to explicitly include the dataframe as an argument
to the `filter()` and `select()` functions any more.

Some may find it helpful to read the pipe like the word "then". For instance,
in the above example, we take the dataframe `kings`, *then* we `filter`
for rows with `House == "Gorm"`, *then* we `select` columns `Name:End_year`.
The **`dplyr`** functions by themselves are somewhat simple,
but by combining them into linear workflows with the pipe, we can accomplish
more complex data wrangling operations.

If we want to create a new object with this smaller version of the data, we
can assign it a new name:

```{r, purl=FALSE}
kings_gorm <- kings %>%
    filter(House == "Gorm") %>%
    select(Name:End_year)

kings_gorm
```

Note that the final dataframe (`kings_gorm`) is the leftmost part of this
expression.

### Exercise 1

Using pipes, subset the `kings` data to include kings with non-missing reign dates, who were born in the 14th century
(`Birth_year` > 1300 & `Birth_year` < 1399) and retain only the columns `Name`,
`House`,  `Start_year` and `End_year`.

```{r, solution = TRUE}
kings %>%
    filter(Birth_year > 1300 &  Birth_year < 1399) %>%
    select(Name, House, Start_year, End_year) 
```


## Mutate

Frequently you'll want to create new columns based on the values in existing
columns, for example to do unit conversions, or to find the ratio of values in
two columns. For this we'll use `mutate()`.

We might be interested in the number of years each king lived, reigned, and what the mid-year of each ruler's reign was (such as you will need to plot the rulers on a timeline):

```{r, purl=FALSE}
kings %>%
  mutate(Reign_duration = End_year - Start_year)
  
```

We may be interested in investigating whether being a member of a specific House had any effect on the duration of reign. To look at this relationship, we will first remove
data from our dataset where the dates are missing.
These cases are recorded as "NULL" in the dataset.

To remove these cases, we could insert a `filter()` in the chain:

```{r, purl=FALSE}
kings %>%
    filter(!is.na(Start_year)) %>%
    mutate(Reign_duration = End_year - Start_year) %>% 
    mutate(Midyear = End_year - (End_year-Start_year) / 2 )
```

The `!` symbol negates the result of the `is.na()` function. Thus, if `is.na()`
returns a value of `TRUE` (because the `Start_year` is missing), the `!` symbol
negates this and says we only want values of `FALSE`, where `Start_year` **is
not** missing.


### Exercise 2

Load the `kings` data and create a new dataframe that meets the following
criteria: contains the `Name` and `House` column and three new columns called
`Years_lived`,`Reign_duration`, and `Midyear` containing values that encompass the monarch's life, duration of reign and the middle year of their reign (end of rule year - duration/2).
Only the rows where `Reign_duration` is greater than 25 should be shown in the
final dataframe.

**Hint**: think about how the commands should be ordered to produce this data
frame!

```{r, solution = TRUE}
kings <- kings %>%
    mutate(Years_lived = Death_year - Start_year) %>% 
    mutate(Reign_duration = End_year - Start_year) %>% 
    mutate(Midyear = End_year - (End_year-Start_year) / 2 ) %>% 
    select(Name, House, Gender, Years_lived, Reign_duration, Midyear)
```


## Split-apply-combine data analysis and the summarize() function

Many data analysis tasks can be approached using the *split-apply-combine*
paradigm: split the data into groups, apply some analysis to each group, and
then combine the results. **`dplyr`** makes this very easy through the use of
the `group_by()` function.

### The `summarize()` function

`group_by()` is often used together with `summarize()`, which collapses each
group into a single-row summary of that group.  `group_by()` takes as arguments
the column names that contain the **categorical** variables for which you want
to calculate the summary statistics. So to compute the average .........:

```{r, purl=FALSE}
kings %>%
    group_by(House) %>%
    summarize(mean_rule = mean(Reign_duration))
```

You may also have noticed that the output from these calls doesn't run off the
screen anymore. It's one of the advantages of `tbl_df` over dataframe.


You can also group by multiple columns:

```{r, purl=FALSE}
kings %>%
    group_by(House, Gender) %>%
    summarize(mean_rule = mean(Reign_duration))
```

When grouping both by `House` and `Gender`, we see rows in our table for
monarchs of different gender. 


### Practice


1. Try summarizing the kings by the average number of years they lived, as an exercise.


```{r, purl=FALSE}
kings %>%
    group_by(House) %>%
    summarize(mean_years = mean(Years_lived))
```

Notice that there are a number of NAs in the result. We can exclude those data from our table using a filter step.


```{r, purl=FALSE, solution = TRUE}
kings %>%
    filter(!is.na(Years_lived)) %>%
    group_by(House) %>%
    summarize(mean_years = mean(Years_lived))
```


2. Can you round the mean_years so we have only 1 decimal?

```{r, solution = TRUE}
kings %>%
    filter(!is.na(Years_lived)) %>%
    group_by(House) %>%
    summarize(mean_years = round(mean(Years_lived),1))
```


Once the data are grouped, you can also summarize multiple variables at the same
time (and not necessarily on the same variable). For instance, we could add a
column indicating the minimum and maximum years ruled for each House for each group
(men or women):

```{r, purl=FALSE}
kings %>%
    filter(!is.na(Reign_duration)) %>%
    group_by(House, Gender) %>%
    summarize(max_rule = max(Reign_duration),
              min_rule = min(Reign_duration))
```

It is sometimes useful to rearrange the result of a query to inspect the values.
For instance, we can sort on `min_membrs` to put the group with the shortest reign first:

```{r, purl=FALSE}
kings %>%
    filter(!is.na(Reign_duration)) %>%
    group_by(House, Gender) %>%
    summarize(max_rule = max(Reign_duration),
              min_rule = min(Reign_duration)) %>%
    arrange(min_rule)
```

To sort in descending order, we need to add the `desc()` function. If we want to
sort the results by decreasing order of minimum household size:

```{r, purl=FALSE}
kings %>%
    filter(!is.na(Reign_duration)) %>%
    group_by(House, Gender) %>%
    summarize(max_rule = max(Reign_duration),
              min_rule = min(Reign_duration)) %>%
    arrange(desc(min_rule))
```

### Counting

When working with data, we often want to know the number of observations found
for each factor or combination of factors. For this task, **`dplyr`** provides
`count()`. For example, if we wanted to count the number of rows of data for
each House, we would do:

```{r, purl=FALSE}
kings %>%
    count(House)
```

For convenience, `count()` provides the `sort` argument to get results in
decreasing order:

```{r, purl=FALSE}
kings %>%
    count(House, sort = TRUE)
```


## Saving data 
Now that you have nicely enriched the kings, we can save this dataframe to our `data_output` directory.

```{r, purl=FALSE, eval=FALSE}
write_csv(kings, file = "data_output/kings_plotting.csv")
```

```{r, purl=FALSE, eval=TRUE, echo=FALSE}
if (!dir.exists("data_output")) dir.create("data_output")
write_csv(kings, "data_output/kings_plotting.csv")
```

### Exercise 3

1. How many Danish monarchs enjoyed a longer-than-average reign?

Suggested steps:

* Select the `Names` and `Reign_duration` columns with `select()`
* `filter()` the kings dataset `Reign_duration` > the-average-length that you calculated before and count the resulting rows with `count()`

```{r, solution = TRUE}
# for the average years of rule
kings %>%
  
  filter(Reign_duration > 29.5) %>% 
  count()
```



* Use `mutate()` to create a new column `AboveAverage` and populate it with TRUE/FALSE which will be a response to the question of whether the `Reign_duration` value is bigger or smaller than average-length
* Group the unique values in the `AboveAverage` column with `group_by()` and then `count()`. What is your result?


```{r, solution = TRUE}
kings %>%
  mutate(AboveAverage = Reign_duration > 29) %>% 
  group_by(AboveAverage) %>% 
  count()
```



2. Sort kings by reign duration in ascending and descending order. Which three ruled the longest/shortest?

3. How many days did the three longest-ruling kings rule?  Use mutate() to create a column called Days calculating the total umber of days they ruled. Consider transition year in your calculation! (366 days)


```{r, solution = TRUE}
kings %>%
  arrange(Reign_duration)
```


4. What is the total number of Christians and Frederiks among the Danish kings? Search on patterns among character strings can be done with `grepl()` or `str_detect()` functions and a wild card.


```{r, solution = TRUE}
kings %>%
  filter(str_detect(Name, "Christian*|Fred*"))
```


```{block, opts.label = "information"}
- Use the `dplyr` package to manipulate dataframes.
- Use `select()` to choose variables from a dataframe.
- Use `filter()` to choose data based on values.
- Use `group_by()` and `summarize()` to work with subsets of data.
- Use `mutate()` to create new variables.
```

