---
title: Data Visualisation with ggplot2
teaching: 80
exercises: 35
output:
  unilur::tutorial_html_solution:
    toc: true
    toc_float: true
    toc_depth: 3
    suffix: ""
    theme: journal
    highlight: kate
    number_sections: no
    number_subsections: no
---

```{r knitr_init, echo=FALSE, cache=FALSE, include=FALSE}
library(knitr)
library(unilur)

## Global options
options(max.print="90")
opts_chunk$set(echo=TRUE,
               cache=FALSE, #TRUE
               prompt=FALSE,
               tidy=FALSE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=90)
options(width = 90)

# boxes custom
#remotes::install_github("koncina/unilur")

knitr::opts_template$set(alert = list(box.title = "Watch out!",
                                      box.body = list(fill = "#ffe0d9", colour = "black"),
                                      box.header = list(fill = "#FFAD99", colour = "black"),
                                      box.collapse = NULL,
                                      box.icon = "fa-exclamation-triangle"))
knitr::opts_template$set(solution = list(box.title = "Solution",
                                         box.body = list(fill = "#e6f6e7", colour = "black"),
                                         box.header = list(fill = "#ace1af", colour = "black"),
                                         box.icon = "fa-check-square",
                                         box.collapse = TRUE))
knitr::opts_template$set(objectives = list(box.title = "Objectives",
                                            box.body = list(fill = "#bbe8f4", colour = "black"),
                                            box.header = list(fill = "#64c9e6", colour = "black"),
                                            box.icon = "fa-info-circle",
                                            box.collapse = NULL))
knitr::opts_template$set(questions = list(box.title = "Questions",
                                      box.body = list(fill = "#fff9dc", colour = "black"),
                                      box.header = list(fill = "#ffec8b", colour = "black"),
                                      box.icon = "fa-search",
                                      box.collapse = NULL))
knitr::opts_template$set(information = list(box.title = "Keypoints",
                                       box.body = list(fill = "#fff9dc", colour = "black"),
                                      box.header = list(fill = "#ffec8b", colour = "black"),
                                      box.icon = "fa-search",
                                      box.collapse = NULL))

```

```{block, opts.label = "objectives" }
- Produce scatter plots, boxplots, and barplots using ggplot.
- Set universal plot settings.
- Describe what faceting is and apply faceting in ggplot.
- Modify the aesthetics of an existing ggplot plot (including axis labels and colour).
- Build complex and customized plots from data in a data frame.
```

```{block, opts.label = "questions" }

- What are the components of a ggplot?
- How do I create scatterplots, lineplots, and barplots?
- How can I change the aesthetics (ex. colour, transparency) of my plot?
- How can I create multiple plots at once?
```


We start by loading the required package. **`ggplot2`** is also included in the
**`tidyverse`** package.

```{r load-package, message=FALSE, purl=FALSE}
library(tidyverse)
```

If not still in the workspace, load the data we saved in the previous lesson.

```{r load-data, purl=FALSE}
kings <- read_csv("data_output/kings_plotting.csv")
```

## Plotting with **`ggplot2`**

**`ggplot2`** is a plotting package that makes it simple to create complex plots
from data stored in a data frame. It provides a programmatic interface for
specifying what variables to plot, how they are displayed, and general visual
properties. Therefore, we only need minimal changes if the underlying data
change or if we decide to change from a bar plot to a scatterplot. This helps in
creating publication quality plots with minimal amounts of adjustments and
tweaking.

**`ggplot2`** functions work best with data in the 'long' format, i.e., a column for every
dimension, and a row for every observation. Well-structured data will save you
lots of time when making figures with **`ggplot2`**

ggplot graphics are built step by step by adding new elements. Adding layers in
this fashion allows for extensive flexibility and customization of plots.

Each chart built with ggplot2 must include the following

- Data

- Aesthetic mapping (aes)
  
  - Describes how variables are mapped onto graphical attributes
  - Visual attribute of data including x-y axes, color, fill, shape, and alpha

- Geometric objects (geom)
  
  - Determines how values are rendered graphically, as bars (`geom_bar`), scatterplot (`geom_point`), line (`geom_line`), etc.

Thus, the template for graphic in ggplot2 is:

```
<DATA> %>%
    ggplot(aes(<MAPPINGS>)) +
    <GEOM_FUNCTION>()
```

Remember from the last lesson that the pipe operator `%>%` places the result of the previous line(s) into the first argument of the function. **`ggplot`** is a function that expects a data frame to be the first argument. This allows for us to change from specifying the `data =` argument within the `ggplot` function and instead pipe the data into the function.

- use the `ggplot()` function and bind the plot to a specific data frame.

```{r ggplot-steps-1, eval=FALSE, purl=FALSE}
kings %>%
    ggplot()
```

- define a mapping (using the aesthetic (`aes`) function), by selecting the variables to be plotted and specifying how to present them in the graph, e.g. as x/y positions or characteristics such as size, shape, color, etc.

```{r ggplot-steps-2, eval=FALSE, purl=FALSE}
kings %>%
    ggplot(aes(x = Midyear, y = Reign_duration))
```

- add 'geoms' – graphical representations of the data in the plot (points,
  lines, bars). **`ggplot2`** offers many different geoms; we will use some
  common ones today, including:
  
  - `geom_point()` for scatter plots, dot plots, etc.
  - `geom_boxplot()` for, well, boxplots!
  - `geom_line()` for trend lines, time series, etc.

To add a geom to the plot use the `+` operator. Because we have two continuous variables, let's use `geom_point()` first:

```{r first-ggplot, purl=FALSE}
kings %>% 
  ggplot(aes(x = Midyear, y = Reign_duration)) +
  geom_point()    # basic scatterplot
 
```

The `+` in the **`ggplot2`** package is particularly useful because it allows
you to modify existing `ggplot` objects. This means you can easily set up plot
templates and conveniently explore different types of plots, so the above plot
can also be generated with code like this, similar to the "intermediate steps"
approach in the previous lesson:


```{r first-ggplot-smooth, purl=FALSE}
ggplot(kings, aes(x = Midyear, y = Reign_duration)) +
  geom_point()+    # basic scatterplot
  geom_smooth()   # visual trend
```

```{block, box.title = "Note", box.body = list(fill = "white"), box.icon = "fa-star"}

- Anything you put in the `ggplot()` function can be seen by any geom layers
  that you add (i.e., these are universal plot settings). This includes the x-
  and y-axis mapping you set up in `aes()`.
- You can also specify mappings for a given geom independently of the mapping
  defined globally in the `ggplot()` function.
- The `+` sign used to add new layers must be placed at the end of the line
  containing the *previous* layer. If, instead, the `+` sign is added at the
  beginning of the line containing the new layer, **`ggplot2`** will not add
  the new layer and will return an error message.

```


```{r ggplot-object, eval=FALSE, purl=FALSE}
kings_plot <- ggplot(kings, aes(x = Midyear, y = Reign_duration)) +
  geom_point()+    # basic scatterplot
  geom_smooth()+   # visual trend
  labs(title = "How long danish kings ruled over time", 
       x = "Year ", y = "Year they ruled") +   # better title and axes' labels
  theme_bw() +                                 # cleaner look 
  theme(text = element_text(size = 14))        # bigger font to make readable    
```

```{r ggplot-with-plus-position, eval=FALSE, purl=FALSE}
## This is the correct syntax for adding layers
kings_plot +
     geom_text(aes(label=Name), size=3) 

## This will not add the new layer and will return an error message
kings_plot
+   geom_text(aes(label=Name), size=3) 

```

## Building your plots iteratively

Building plots with **`ggplot2`** is typically an iterative process. We start by
defining the dataset we'll use, lay out the axes, and choose a geom:

```{r create-ggplot-object, fig.alt="Scatter plot of number of items owned versus number of household members.", purl=FALSE}
kings %>%
    ggplot(aes(x = Midyear, y = Reign_duration)) +
    geom_point()
```

Then, we start modifying this plot to extract more information from it. For
instance, when inspecting the plot we notice that points only appear at the
intersection of whole numbers of `Midyear` and `Reign_duration`. 


To colour each village in the plot differently, you could use a vector as an input
to the argument **`color`**.  However, because we are now mapping features of the
data to a colour, instead of setting one colour for all points, the colour of the
points now needs to be set inside a call to the **`aes`** function. When we map
a variable in our data to the colour of the points, **`ggplot2`** will provide a
different colour corresponding to the different values of the variable. We will
continue to specify the value of **`alpha`**, **`width`**, and **`height`**
outside of the **`aes`** function because we are using the same value for
every point. ggplot2 understands both the Commonwealth English and
American English spellings for colour, i.e., you can use either `color`
or `colour`. Here is an example where we color points by the **`village`**
of the observation:

```{r color-by-species, purl=FALSE}
kings %>%
    ggplot(aes(x = Midyear, y = Reign_duration)) +
    geom_point(aes(color = House), width = 0.2, height = 0.2)
```

There appears to be a positive trend between number of household
members and number of items owned (from the list provided). Additionally,
this trend does not appear to be different by village.

:::::::::::::::::::::::::::::::::::::::::  callout

## Notes

As you will learn, there are multiple ways to plot the a relationship
between variables. Another way to plot data with overlapping points is
to use the `geom_count` plotting function. The `geom_count()`  function
makes the size of each point representative of the number of data items
of that type and the legend gives point sizes associated to particular
numbers of items.

```{r color-by-species-notes, fig.alt="Previous plot with dots colored by house.", purl=FALSE}
kings %>%
   ggplot(aes(x = Midyear, y = Reign_duration, color = House)) +
   geom_count()
```


### Exercise 1

Use what you just learned to create a scatter plot of `midpoint of rulers' reign` by `duration of length`
with the `house` showing in different colours. Does this
seem like a good way to display the relationship between these variables?
What other kinds of plots might you use to show this type of data?


```{r, fig.alt="Scatter plot of royal houses by reign of duration.", answer=TRUE, purl=FALSE, solution = TRUE}
kings %>%
    ggplot(aes(x = Midyear, y = Reign_duration)) +
    geom_jitter(aes(color = House),
	  #  alpha = 0.3,
		       height = 0.2)
```

This is not a great way to show this type of data because it is difficult to
distinguish any trend in such a wide spread of dots. What other plot types could help you visualize
this relationship better?

::::::::::::::::::::::::::::::::::::::::::::::::::

## **`ggplot2`** themes

In addition to `theme_bw()`, which changes the plot background to white,
**`ggplot2`** comes with several other themes which can be useful to quickly
change the look of your visualization. The complete list of themes is available
at [https://ggplot2.tidyverse.org/reference/ggtheme.html](https://ggplot2.tidyverse.org/reference/ggtheme.html). `theme_minimal()` and
`theme_light()` are popular, and `theme_void()` can be useful as a starting
point to create a new hand-crafted theme.

The
[ggthemes](https://jrnold.github.io/ggthemes/reference/index.html)
package provides a wide variety of options (including an Excel 2003 theme). The
[**`ggplot2`** extensions website](https://exts.ggplot2.tidyverse.org/) provides a list
of packages that extend the capabilities of **`ggplot2`**, including additional
themes.



### Exercise 2

Experiment with at least two different themes. Build the previous plot
using each of those themes. Which do you like best?


## Customization

Take a look at the [**`ggplot2`** cheat
sheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-visualization.pdf),
and think of ways you could improve the original smoothed `kings_plot`.

Now, let's look at different themes and make sure you have changed names of axes to something more informative than 'Midyear' and
'Reign_duration' and add a title to the figure:

```{r ggplot-customization-2, purl=FALSE}
kings %>%
    ggplot(aes(x = Midyear, y = Reign_duration)) +
    geom_jitter(aes(color = House),
	  #  alpha = 0.3,
		       height = 0.2)+
    theme_classic()  # try also theme_minimal, and others
    
```



### Exercise 1

With all of this information in hand, please take another five minutes to
either improve one of the plots generated in this exercise or create a
beautiful graph of your own. Use the RStudio [**`ggplot2`** cheat sheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-visualization.pdf)
for inspiration. Here are some ideas:

- See if you can make the bars white with black outline.
- Try using a different colour palette (see
  [http://www.cookbook-r.com/Graphs/Colors\_(ggplot2)/](https://www.cookbook-r.com/Graphs/Colors_\(ggplot2\)/)).



After creating your plot, you can save it to a file in your favourite format. The Export tab in the **Plot** pane in RStudio will save your plots at low resolution, which will not be accepted by many journals and will not scale well for posters.

Instead, use the `ggsave()` function, which allows you to easily change the dimension and resolution of your plot by adjusting the appropriate arguments (`width`, `height` and `dpi`).

Make sure you have the `fig_output/` folder in your working directory.

```{r ggsave-example, eval=FALSE, purl=FALSE}
my_plot <- 

ggsave("fig_output/name_of_file.png", my_plot, width = 15, height = 10)
```

Note: The parameters `width` and `height` also determine the font size in the saved plot.

```{block, opts.label = "information"}

- `ggplot2` is a flexible and useful tool for creating plots in R.
- The data set and coordinate system can be defined using the `ggplot` function.
- Additional layers, including geoms, are added using the `+` operator.
- Boxplots are useful for visualizing the distribution of a continuous variable.
- Barplots are useful for visualizing categorical data.
- Faceting allows you to generate multiple plots based on a categorical variable.
```

